% !TeX root = ../dissertation.tex
\chapter{Evaluation}

\section{Correctness}

The initial compiler fully covered the entire instruction set

Nearly all tests passed except expected failures

The second compiler could not do everything the initial compiler could but this was ok because it
failed gracefully with fallback to the initial compiler.

\section{The benchmark suite - Sandmark}

Project by OCaml labs here at the CL. For multicore ocaml but good single core too. Most complete
effort (mention previous effort used by flambda). Mention OPAM/dune tooling.

Explain criteria for selection of bytecode. Explain implicit bias by longer running programs.

\section{Benchmark results}

Show impressive graph with summary of all results

\subsection{Initial compiler}

Show table of some results and graph of speedups. Mention distinction between tracing code + it
disabled.

\subsection{Overhead of supporting recompilation}

Show table/graph showing performance loss by extra indirection

\subsection{Optimsed compiler}

Show table graph

\section{Evaluation of results and design trade-offs}

Write some stuff about when to use which compiler. Explain why certain ones were good/bad.

Quantify overhead for shorter running programs and explain it's likely to be slower. Use
self-hosting the compiler as a particularly bad case for the system.

Go into better cases for it and explain why it works. Possibly include code snippets.

\section{Summary}

Reiterate 100\% correctness. Reiterate performance numbers.